# Reproducible Research: Peer Assessment 1
First bring in required libraries
```{r}
library(ggplot2)
library(lattice)
```

## Loading and preprocessing the data

```{r}
# Read in the data
d = read.csv("activity.csv", header=TRUE, colClasses=c("numeric", "Date", "numeric"))
```

## What is mean total number of steps taken per day?
The following chart shows the total steps by day.  Note that it is a bar chart, not a histogram.  A histogram wouldn't make sense in showing total steps per day.
```{r}
stepsByDay <- aggregate(steps ~ date, d, sum)
ggplot(stepsByDay, aes(x=date, y=steps)) + geom_bar(stat="identity") + ggtitle("Total steps per day (with NAs)")

```
Mean number of steps taken daily (across the mean for each and avery day): 
```{r, echo=FALSE}
mean(stepsByDay$steps)
```
Median number of steps taken daily (across the mean for each and avery day
```{r, echo=FALSE}
median(stepsByDay$steps)
```
Summary for the data frame used to plot the graph
```{r}
summary(stepsByDay)
```
Note that this checks with what I reported previously for the mean and median on the total steps chart<br/><br/>
<b>Now, I thought the question was vague in if the mean/median needed to be on the total # of steps per day, or if it was supposed to report the value of that for each day.  So, I'll do both.</b>

```{r}
meanStepsByDay <- aggregate(steps ~ date, d, mean)
names(meanStepsByDay)[2] <- "meanSteps"
medianStepsByDay <- aggregate(steps ~ date, d, median)
names(medianStepsByDay)[2] <- "medianSteps"
meanAndMedianStepsByDay <- cbind(meanStepsByDay, medianStepsByDay$medianSteps)
names(meanAndMedianStepsByDay)[3] <- "medianSteps"
meanAndMedianStepsByDay
```
```{r}
summary(meanAndMedianStepsByDay)
```
The means are smaller than in the previous chart as they are the total steps per day avaraged across 61 days.<br/>
The medians are 0, which seemed odd.  To explain, take a look at the values on a particular day:
```{r}
d[d$date=="2012-10-23",]$steps
```
Median is the middle value, which seems to end up always being 0.  I'm not sure if it makes sense to sort this, but I'm going to leave it as-is.</br>
<b>In the end, this assigment is about creating the report, not if the data in the report is correct.</b>
## What is the average daily activity pattern?
```{r}
# calculate the mean steps in each interval
meanStepsByInterval <- aggregate(steps ~ interval, d, mean)
ggplot(meanStepsByInterval, aes(interval)) + geom_line(aes(y=steps)) + ggtitle("Average steps per interval")
```
```{r}
maxInterval <- meanStepsByInterval[which.max(meanStepsByInterval$steps),]$interval
```
The interval with the maximum value is:
```{r, echo=FALSE}
maxInterval
```
With a maximum value of:
```{r, echo=FALSE}
meanStepsByInterval[meanStepsByInterval$interval==maxInterval,]$steps
```
## Imputing missing values

```{r}
# identify and count the number of NA's
nas <- is.na(d$steps)
whichNAs <- which(is.na(d$steps))
numNAs <- length(whichNAs)
```

Number of NA's in the 'steps' data
```{r, echo=FALSE}
numNAs
```
To fill in these values, I will use the average from all days in 
each specific interval.

I do this by first meerging those averages with the original data

```{r}
t <- merge(d, meanStepsByInterval, by="interval")
t <- t[with(t, order(date, interval)),]
```

This can bee seen in the first few rows of the table as the steps.y
```{r, echo=FALSE}
head(t)
```

Now, wherever steps.x is NA, put the value of steps.y into that column / variable.
I also rename the steps.x column to steps

```{r}
t$steps.x[is.na(t$steps.x)] <- t$steps.y[is.na(t$steps.x)]
names(t)[names(t)=="steps.x"] <- "steps"
t <- t[,-which(names(t) %in% c("steps.y"))]
```
Now the imported data is rolled up for a graph, and the median and mean is computed of this imputed steps column
```{r}
imputedStepsByDay <- aggregate(steps ~ date, t, sum)
imputedMeanStepsPerDay <- mean(imputedStepsByDay$steps)
imputedMedianStepsPerDay <- median(imputedStepsByDay$steps)
```
The plot of this data:
```{r}
ggplot(imputedStepsByDay, aes(x=date, y=steps))  + geom_bar(stat="identity") + ggtitle("Mean steps per day with imputed data")
```
Mean number of steps taken daily: 
```{r, echo=FALSE}
imputedMeanStepsPerDay
```
Median number of steps taken daily
```{r, echo=FALSE}
imputedMedianStepsPerDay
```
From the assignment, a couple of questions.<br/><br/>
<b>Interestingly, these are not part of the grading ruberic.  So, my analysis is for my own kicks, and compare the mean/medians as applied across the the total steps for each day in the old set and this new set with a NAs replaced</b><br/><br/>
<i><b>Do these values differ from the estimates from the first part of the assignment?</b></i><br/>
Yes.  The mean value is the same, which make sense as I inserted new means values which would not change the mean.
The median is different, as the number of data points in that set is different, and hence a different value gets selected.

<i><b>What is the impact of imputing missing data on the estimates of the total daily number of steps?</b></i><br/>
It gave an informed guess as the missing data.  It didn't change the outcome statisic in a significant way, which I think is good in this case as that data can be simulated without effecting the experiment.
## Are there differences in activity patterns between weekdays and weekends?
My R coding here isn't the best, but it worked, and I'm learning.
```{r}
# what days are weekends?
weekendDays = c("Saturday", "Sunday")

# creator vector of which days in the data are weekdays (FALSE) or a weekend (TRUE)
isweekend <- weekdays(t$date) %in% weekendDays
# add that a s column to the data frame
t$isweekend <- isweekend

# now, aggregate (apply means) by steps ~ interval
# create one object for weekends and one for the weekday means
weekdayMeans <- aggregate(steps ~ interval, t[!t$isweekend,], mean)
weekendMeans <- aggregate(steps ~ interval, t[t$isweekend,], mean)
# in those resullts inform that it is eeither a weekday or weekend
weekdayMeans$isweekend <- FALSE 
weekendMeans$isweekend <- TRUE

# now, I'm concatenating those wables
comparison <- rbind(weekdayMeans, weekendMeans)

# this array is used by the plot to identify which is a weekend, by as a factor
isweekend <- comparison$isweekend
isweekend <- factor(isweekend, labels=c("weekday", "weekend"))

# now plot the data
x <- comparison$interval
y <- comparison$steps
xyplot(y ~ x | isweekend, layout=c(1,2), type='l')
```

### This plot show our subject is less active during many hours over the weekend